Optimal Neural Summarisation for Full-Field Implicit Inference by Density Estimation

### 2023-10-27 paper/paper.tex
change disp sections
27/10

%-------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{Introduction}
Weak gravitational lensing by the large-scale structure is caused by the presence of foreground matter bending the light emitted by background galaxies. Being sensitive to the large-scale structure of the universe, it is one of the most promising tools for investigating the nature of dark energy, the origin of the accelerating expansion of the universe and estimating cosmological parameters. Future cosmological surveys, such as the Legacy Survey of Space and Time (LSST) of the Vera C. Rubin Observatory \citep{ivezic2019lsst}, the Nancy Grace Roman Space Telescope \citep{spergel2015wide}, and the Euclid Mission \citep{laureijs2011euclid}, will rely on weak gravitational lensing as one of the principal physical probes to address unresolved questions in current cosmology.
 As these surveys become deeper, they will be able to access more non-Gaussian features of the matter fields. This makes the standard weak lensing analyses, which rely on two-point statistics such as the 2-point shear correlation or the angular power spectrum, sub-optimal. These analyses are unable to fully capture the non-Gaussian information imprinted in the lensing signal and can only access Gaussian information.

To overcome this limitation, several higher-order statistics have been introduced. These include weak lensing peak counts \citep{liu2015cosmology,  liu2015cosmological, lin2015new, kacprzak2016cosmology, peel2017cosmological, shan2018kids, martinet2018kids, ajani2020constraining, harnois2021cosmic, zurcher2022dark}, wavelet and scattering transform \citep{ajani2021starlet, cheng2021weak}, the one point PDF \citep{liu2019constraining, uhlemann2020fisher, boyle2021nuw}, Minkowski functionals \citep{kratochvil2012probing, petri2013cosmology}, moments of mass maps \citep{gatti2021dark},  and 3 point statistics \citep{takada2004cosmological, semboloni2011weak, rizzato2019tomographic, halder2021integrated}. 
Although these methods have proven to improve cosmological constraints, they rely on summary statistics, for which an analytical description is often lacking. These methods frequently assume a Gaussian likelihood and require accurate computation of the covariance matrix, which can be challenging.

@@

In this paper, we present a comparison of the performance of various neural network-based compression schemes within the Likelihood-Free Inference (LFI) framework. These methods differ in terms of the loss functions used to train the neural network, but they share the same inference strategy based on neural density estimation. To address the need for result validation in the LFI context, we employ a Bayesian forward modeling approach that enables inference on the joint posterior of the cosmological parameters and the convergence field. The forward model is based on \href{https://github.com/DifferentiableUniverseInitiative/sbi_lens}{\url{SBILens}}, a Python package designed for weak lensing inference with differentiable simulator implemented in Jax. \texttt{SBILens} allows the sampling of convergence maps in a tomographic setting, accounting for the cross-correlation between different redshift bins.

The paper is structured as follows: in \autoref{Sec:Motivation}, we illustrate the motivation behind this work. In \autoref{Sec:the SBILens framework}, we introduce the \texttt{SBILens} framework and describe the simulated data used in this work. In \autoref{Sec:experiment}, we detail the inference strategy and the three different approaches we used: the power spectrum, the map-based inference based on the Bayesian forward-modeling, and the map-based inference based on the LFI. In the same section, we also provide a detailed overview of different neural compression strategies. In \autoref{Sec:results}, we discuss the results and validate the LFI approaches. Finally, we conclude in \autoref{Sec:conclusion}. 
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{Motivation}\label{Sec:Motivation}
%--------------------------------------------------------------------
With the increased statistical power of stage IV surveys, our cosmological analysis should not rely on the measurement of sub-optimal summary statistics, which may not fully capture the non-Gaussian information present in the lensing field at the scales accessible to future surveys. In this paper, we introduce a forward model that directly extracts information from the convergence maps, rather than relying on the analytical evaluation of summary statistics. By doing so, we aim to preserve all available information and facilitate the incorporation of systematic effects and the combination of multiple cosmological probes through joint simulations.
In a forward model context, the simulator of the observables serves as our physical model, where each component is tractable. These models, often referred to as \textit{probabilistic program}, as illustrated by \citet{cranmer2020frontier}, can be described as follows: The models take as input a vector parameter $\bm{\theta}$, which describes the underlying deterministic model. Then, they sample internal states $\bm{z}$, dubbed \textit{latent variables}, from the distribution $p(\bm{z}|\bm \theta)$. These states can be directly or indirectly related to a physically meaningful state of the system. Finally, the models generate the output $\bm x$ from the distribution $p(\bm x|\bm \theta,\bm{z})$, where $\bm{x}$ represents the observations.

The ultimate goal of Bayesian inference in cosmology is to compute the posterior distribution:
\begin{equation}\label{Eq:posterior}
     p(\bm{\theta}|\bm{x})= 
     \frac{p(\bm{x}|\bm{\theta})p(\bm{\theta})}
     {\int d\bm{\theta'}p(\bm{x}|\bm{\theta}')p(\bm{\theta}')},
\end{equation}
however, a problems arises because the marginal likelihood $p(\bm{x}|\bm{\theta})$ is typically intractable:
\begin{equation}
    p(\bm{x}|\bm{\theta})=\int p(\bm{x},\bm{z}|\bm{\theta}) d\bm{z}=\int p(\bm{x}|\bm{z},\bm{\theta})p(\bm{z}|\bm{\theta}) d\bm{z}.
\end{equation}
since it involves integrating over all potential paths through the latent space.
To overcome this limitation while still capturing the full information content of the data, two different approaches have been proposed in the literature. Although these approaches are often referred to by different names, hereinafter we will make the following distinction:
%--------------------------------------------------------------------
\paragraph{\textbf{Explicit inference}} referring to all Likelihood-based inference approaches.  
This approach involves treating the simulator as a probabilistic model and performing inference over the joint posterior:
\begin{equation}
        p(\bm{\theta},\bm{z}|\bm{x})\propto  p(\bm{x},\bm{z}|\bm{\theta}) p(\bm{z}|\bm{\theta}).
\end{equation}
In the context of the forward model and full-field analysis, Bayesian Hierarchical Models (BHMs) fall into this category.
A Bayesian forward approach involves using a given physical model to predict observations and then comparing these predictions with observations to infer the parameters of the model.
However, performing a global analysis on complex and large datasets, such as a lensing dataset, is not feasible. Therefore, the idea of dividing the global analysis into several subgroups, analyzing them in multiple steps, and using the output of each of those as input for the next one. From here, the \textit{hierarchical} nature of this methodology.
%--------------------------------------------------------------------
\paragraph{\textbf{Implicit inference}} referring to all the approaches that try to infer the marginal likelihood from simulations. This second class of approaches involves treating the simulator as a black box with only the ability to sample from the joint distribution:
\begin{equation}
    (\bm{x}, \bm{\theta})\sim p(\bm{x}, \bm{\theta}).
\end{equation}
Within this class of methods, we can differentiate between more traditional methods such as Approximate Bayesian Computation (ABC) and Density Estimation Likelihood-Free Inference (DELFI) methods. ABC employs rejection-criteria based approaches to approximate the likelihood by comparing simulations with data. 
In this work, our focus will be on DELFI methods, which approach the inference task as a density estimation problem. \\
In particular, the standard deep learning approach for Likelihood-Free Inference can be described as separated into two distinct steps:
\begin{enumerate}
    \item  Automatically learning an optimal low-dimensional summary statistic.
    \item Using Neural Density Estimation in low dimensions to infer the target distributions.
\end{enumerate}
In the first step, we introduce a parametric function $F_{\varphi}$ such that:
     \begin{equation}
         \bm{t}=F_{\varphi}(\bm{\theta}),
     \end{equation}
which aims to reduce the dimensionality of the data while preserving information. Typically, the compressed statistics $\bm{t}$ is assumed to have the same dimension of $\bm{\theta}$. \\
In the second step, Neural Density Estimation can target either building an estimate $p_{\varphi}$ of the likelihood function $p(\bm{x}|\bm{\theta})$ (referred to as the Neural Likelihood Estimation (NLE) task), or targeting the posterior distribution $p(\bm{\theta}|\bm{x})$, (known as Neural Posterior Estimation (NPE) task).

@@

Therefore, our primary focus will be to investigate the effectiveness and validity of these compression techniques in the context of weak lensing applications (within the limits of our setting).

@@

%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{The SBILens framework}\label{Sec:the SBILens framework}
To investigate the questions above, we have developed the Python package \href{https://github.com/DifferentiableUniverseInitiative/sbi_lens}{\texttt{SBILens}}, a weak lensing differentiable simulator based on a log-normal model. \texttt{SBILens} enables the sampling of convergence maps in a tomographic setting while considering the cross-correlation between different redshift bins.
\subsection{Lognormal Modeling}\label{Sec:Lognormal Modeling}
For various cosmological applications, the non-Gaussian field can be modeled as a Lognormal field \citep{coles1991lognormal,bohm2017bayesian}.
This model offers the advantage of generating the matter or convergence field rapidly while allowing the extraction of information beyond the two-point statistics. 
Although studies demonstrated that this model fails in describing the 3D field \citep{klypin2018density}, it properly describes the 2D convergence field \citep{clerkin2017testing, xavier2016improving}.
Assuming a simulated Gaussian convergence map $\kappa_g$, whose statistical properties are fully described by its power spectrum $C_{\ell}$ we know that this model is not a suitable representation of late-time and more evolved structures. One potential solution is to find a transformation $f(\kappa_g)$ of this map that captures the non-Gaussian features in the convergence field. In doing so, it is crucial to ensure that the transformed map maintains the correct mean and variance, effectively recovering the correct two-point statistics.
Denoting $\mu$ and $\sigma_g^2$ the mean and covariance matrix of $\kappa_g$ respectively, we can define the transformed convergence $\kappa_{ln}$ as a shifted lognormal random field:
\begin{equation}\label{Eq:log_norm_kappa}
    \kappa_{ln}=e^{\kappa_{g}}-\lambda, 
\end{equation}
where $\lambda$ is a free parameter that determines the shift of the lognormal distribution. The convergence $\kappa$ in a given redshift bin is fully determined by the shift parameter $\lambda$, the mean $\mu$ of the associated Gaussian field $\kappa_g$, and its variance $\sigma_{g}^2$.
The correlation of the lognormal field, denoted as $\xi_{ln}$, is also a function of these variables and is related to $\xi^{ij}_g$ through the following equations:
\begin{align}
    \xi^{ij}_{ln}(\theta) & \equiv \lambda_i \lambda_j (e^{ \xi^{ij}_g(\theta)}-1) \nonumber \\ 
    \xi^{ij}_g(\theta)&=\log{\left[ \frac{\xi^{ij}_{ln}(\theta)}{\lambda_i \lambda_j}+1\right ]}. \label{Eq:log_norm_corr}
\end{align}
Here $i$ and $j$ define a pair of redshift bins.
The parameter $\lambda$, also known as \textit{minimum convergence
parameter}, defines the lowest values for all possible values of $\kappa$.
The modeling of the shift parameter can be approached in various ways. For example, it can be determined by matching moments of the distribution \citep{xavier2016improving} or by treating it as a free parameter \citep{hilbert2011cosmic}. In general, the value of $\lambda$ depends on the redshift, cosmology, and the scale of the field at which smoothing is applied.

While it is straightforward to simulate a single map, if we want to constrain the convergence map in different redshift bins, an additional condition must be met. The covariance of the map should recover the correct angular power spectrum:
\begin{equation}\label{power_spectrum_definition}
    \left \langle \tilde{\kappa}^{(i)}_{ln} (\ell)\tilde{\kappa}^{(j)}_{ln}(\ell')\right \rangle =C^{ij}_{ln}(\ell)\delta^{K}(\ell-\ell')
\end{equation}
where $ C^{ij}_{ln}(\ell)$ is the power spectrum of $\kappa_{ln}$ in Fourier space, defined as:
\begin{equation}\label{Eq:log_norm_cls}
    C^{ij}_{ln}(\ell)=2\pi \int_0^{\pi} d\theta \sin{\theta}P_{\ell}(\cos{\theta})\xi^{ij}_{ln}(\theta)
\end{equation}
and $P_{\ell}$ is the Legendre polynomial of order $\ell$. 
Using the lognormal model, we can simultaneously constrain the convergence field in different redshift bins while considering the correlation between the bins, as described by \autoref{Eq:log_norm_corr}.

@@

Each map is reproduced on a regular grid with dimensions of $256 \times 256$ pixels and covers an area of $10\times 10$ deg$^2$.  An example of a tomographic convergence map simulated using the \texttt{SBILens} package is shown in \autoref{fig:convergence_maps}. \denise{TODO: change disp images}

@@

    \includegraphics[width=\textwidth]{figures/bin1and2.pdf}
    \includegraphics[width=\textwidth]{figures/bin3.pdf}
    \includegraphics[width=0.5\textwidth]{figures/bin5.pdf}

@@

     Example of convergence map simulated using the \texttt{SBILens} package.

@@

    Convergence power spectra for different tomographic bin combinations. The solid yellow line shows the measurement from 20 simulated maps using the survey setting described in \autoref{Sec:the SBILens framework}, while the black dashed line shows the theoretical predictions computed using jax-cosmo. In this figure, the shaded regions represent the standard deviation from 20 independent map realizations.

@@

Additionally, we conduct a power spectrum study. Indeed, as discussed in \autoref{Sec:Lognormal Modeling}, log-normal fields offer the advantage of rapidly generating convergent fields while accounting for non-Gaussianities. To emphasize this claim further, along with the full-field analysis, we include a power spectrum analysis. This analysis demonstrates that there is indeed a gain of information when adopting a full-field approach. \\

@@

To construct the explicit map-based inference strategy in the Hierarchical Bayesian framework, we built a likelihood based on the data model described in \autoref{Sec:Lognormal Modeling}. 

@@

However, it is important to note that this approach does not guarantee the recovery of maximally informative summary statistics. Indeed, minimizing the $L_{2}$ norm is equivalent to training the model to estimate the mean of the posterior distribution.  We prove this statement in \autoref{Sec:appendix_Mean Square Error}.

@@

In \autoref{Sec:appendix_Maximum Absolute Error}, we demonstrate that minimizing this loss function is equivalent to training the model to estimate the median of the posterior distribution. While extensively employed in various previous studies \citep{2018PhRvD..97j3515G, fluri2018cosmological, ribli2019weak}, similar to the previous consideration for the MSE, it is important to note that the effectiveness of this loss function in extracting sufficient statistics needs to be demonstrated.

@@

An application of this method in the context of weak lensing can be traced back to the work of \citet{fluri2021cosmological, fluri2022full}. Following \citet{charnock2018automatic}, they presented a compression approach that relies on the optimization of the Cramér-Rao bound. However, their approach does not assume that the summary follows a Gaussian likelihood. The loss function takes the form:
\begin{equation}\label{eq:loss_gfim}
\mathcal{L}=\log{\text{det}(\bm{\Sigma}_{\bm{\theta}}(\bm{t})})-2\log{\left|\text{det}\left(\frac{\partial \Psi_{\bm{\theta}}(\bm{t})}{\partial \bm{\theta}}\right)\right|},
\end{equation}
where $\bm{\Sigma}$ is the covariance matrix, and $\Psi_{\bm{\theta}}(\bm{t})=\mathbb{E}_{p(\bm {x}|\bm {\theta})}[\bm{t}]$.
This loss function is equivalent to the log-determinant of the inverse Fisher matrix used in \autoref{eq:loss_imnn}. \\
% \LM{TODO: condensed introduction; emphasize learning compression at single point; network learns score function which generalises smoothly away from fiducial \cite{Makinen_2021, Makinen_2022, Charnock_2018}, Wandelt (in prep.) 
% }
%--------------------------------------------------------------------
\subsection{Compression strategy}
%--------------------------------------------------------------------
The different compression strategies differ in the loss function employed during the training phase. For all of them, we use the same convolutional neural network architecture to build the compressor, specifically the ResNet-18, a deep ResNet architecture \citep{he2016deep}. The ResNet-18 is implemented using Haiku \citep{haiku2020github}, a Python deep learning library built on top of Jax.

@@

\subsubsection{Inference strategy}\label{Sec:Inference_strategy}

@@

        \caption{Table summarizing the different neural compression schemes used for weak lensing applications. The table includes all studies conducted within the context of implicit analysis (Likelihood-free) and standard Likelihood-based analysis. \\

@@

We present the results of the constraints on the full $\Lambda$CDM parameter space expected for a survey like LSST-Y10. The results are obtained using the simulation procedure outlined in \autoref{Sec:the SBILens framework} and the parameter inference strategy described in \autoref{Sec:Inference_strategy}.

@@

We now compare the constraining power of the three different approaches described in \autoref{Sec:experiment}: the standard two-point statistics and two map-based approaches, the explicit inference and the implicit inference strategy. 

@@

We note that the results obtained using different summaries are generally in agreement with each other, and there are no tensions present for any of the cosmological parameters. We report the marginalized summary constraints in \autoref{tab:summaries}. The results concern the cosmological parameters that are better constrained from weak lensing: $\Omega_c$, $\sigma_8$, $w_0$.

@@

        Constraints on the $\Lambda$CDM parameter space as found in the LSST Y10 survey setup. The constraints are obtained by applying the $C_{\ell}$ (violet contours), the full field explicit inference (orange contours), and the full field implicit inference strategy (blue dashed contours), described in \autoref{Sec:experiment}.

@@

        Constraints on the $\Lambda$CDM parameter space as found in the LSST Y10 survey setup. The constraints are obtained from three CNN map compressed statistics: the MSE (black dashed contours), the MAE (magenta dashed contours), and VMIM (blue contours), described in \autoref{Sec:experiment}.

@@

            $\Omega_c-  \sigma_8$ & 1222 & 2520 & 2526 & 2043 & 2316\\
            $\Omega_c-  w_0$      & 100  & 198  & 190  & 152  & 162\\
            $\sigma_8-  w_0$      & 77   & 176  & 171  & 139  & 149\\

>>>>

%-------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{Introduction}
Weak gravitational lensing by large-scale structures (LSS) is caused by the presence of foreground matter bending the light emitted by background galaxies. Being sensitive to the large-scale structure of the universe, it is one of the most promising tools for investigating the nature of dark energy, the origin of the accelerating expansion of the universe and estimating cosmological parameters. Future cosmological surveys, such as the Legacy Survey of Space and Time (LSST) of the Vera C. Rubin Observatory \citep{ivezic2019lsst}, Nancy Grace Roman Space Telescope \citep{spergel2015wide}, and the Euclid Mission \citep{laureijs2011euclid}, will rely on weak gravitational lensing as one of the principal physical probes to address unresolved questions in current cosmology.
 As these surveys become deeper, they will be able to access more non-Gaussian features of the matter fields. This makes the standard weak-lensing analyses, which rely on two-point statistics such as the 2-point shear correlation or the angular power spectrum, sub-optimal. These analyses are unable to fully capture the non-Gaussian information imprinted in the lensing signal and can only access Gaussian information.

To overcome this limitation, several higher order statistics have been introduced. These include weak-lensing peak counts \citep{liu2015cosmology,  liu2015cosmological, lin2015new, kacprzak2016cosmology, peel2017cosmological, shan2018kids, martinet2018kids, ajani2020constraining, harnois2021cosmic, zurcher2022dark}, wavelet and scattering transform \citep{ajani2021starlet, cheng2021weak}, the one-point probability distribution function \citep[PDF][]{liu2019constraining, uhlemann2020fisher, boyle2021nuw}, Minkowski functionals \citep{kratochvil2012probing, petri2013cosmology}, moments of mass maps \citep{gatti2021dark},  and three-point  point statistics \citep{takada2004cosmological, semboloni2011weak, rizzato2019tomographic, halder2021integrated}. 
Although these methods have proven to improve cosmological constraints, they rely on summary statistics, for which an analytical description is often lacking. These methods frequently assume a Gaussian likelihood and require accurate computation of the covariance matrix, which can be challenging.

@@

In this paper, we present a comparison of the performance of various neural network-based compression schemes within the Likelihood-Free Inference (LFI) framework. These methods differ in terms of the loss functions used to train the neural network, but they share the same inference strategy based on neural density estimation. To address the need for result validation in the LFI context, we employ a Bayesian forward modeling approach that enables inference on the joint posterior of the cosmological parameters and the convergence field. The forward model is based on \href{https://github.com/DifferentiableUniverseInitiative/sbi_lens}{\url{SBILens}}, a Python package designed for weak-lensing inference with differentiable simulator implemented in Jax. \texttt{SBILens} allows the sampling of convergence maps in a tomographic setting, accounting for the cross-correlation between different redshift bins.

The paper is structured as follows: in Section \ref{Sec:Motivation}, we illustrate the motivation behind this work. In Section \ref{Sec:the SBILens framework}, we introduce the \texttt{SBILens} framework and describe the simulated data used in this work. In Section \ref{Sec:experiment}, we detail the inference strategy and the three different approaches we used: the power spectrum, the map-based inference based on the Bayesian forward-modeling, and the map-based inference based on the LFI. In the same section, we also provide a detailed overview of different neural compression strategies. In Section \ref{Sec:results}, we discuss the results and validate the LFI approaches. Finally, we conclude in Section \ref{Sec:conclusion}. 
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{Motivation}\label{Sec:Motivation}
%--------------------------------------------------------------------
With the increased statistical power of stage IV surveys, our cosmological analysis should not rely on the measurement of sub-optimal summary statistics, which may not fully capture the non-Gaussian information present in the lensing field at the scales accessible to future surveys. In this paper, we introduce a forward model that directly extracts information from the convergence maps, rather than relying on the analytical evaluation of summary statistics. By doing so, we aim to preserve all available information and facilitate the incorporation of systematic effects and the combination of multiple cosmological probes through joint simulations.
In a forward model context, the simulator of the observables serves as our physical model, where each component is tractable. These models, often referred to as \textit{probabilistic program}, as illustrated by \citet{cranmer2020frontier}, can be described as follows: The models take as input a vector parameter $\bm{\theta}$, which describes the underlying deterministic model. Then, they sample internal states $\bm{z}$, dubbed \textit{latent variables}, from the distribution $p(\bm{z}|\bm \theta)$. These states can be directly or indirectly related to a physically meaningful state of the system. Finally, the models generate the output $\bm x$ from the distribution $p(\bm x|\bm \theta,\bm{z})$, where $\bm{x}$ represents the observations.

The ultimate goal of Bayesian inference in cosmology is to compute the posterior distribution:
\begin{equation}\label{Eq:posterior}
     p(\bm{\theta}|\bm{x})= 
     \frac{p(\bm{x}|\bm{\theta})p(\bm{\theta})}
     {\int d\bm{\theta'}p(\bm{x}|\bm{\theta}')p(\bm{\theta}')},
\end{equation}
however, a problems arises because the marginal likelihood $p(\bm{x}|\bm{\theta})$ is typically intractable:
\begin{equation}
    p(\bm{x}|\bm{\theta})=\int p(\bm{x},\bm{z}|\bm{\theta}) d\bm{z}=\int p(\bm{x}|\bm{z},\bm{\theta})p(\bm{z}|\bm{\theta}) d\bm{z}.
\end{equation}
since it involves integrating over all potential paths through the latent space.
To overcome this limitation while still capturing the full information content of the data, two different approaches have been proposed in the literature. Although these approaches are often referred to by different names, hereinafter we will make the following distinction:
%--------------------------------------------------------------------
\paragraph{\textbf{Explicit inference}} referring to all Likelihood-based inference approaches.  
This approach involves treating the simulator as a probabilistic model and performing inference over the joint posterior:
\begin{equation}
        p(\bm{\theta},\bm{z}|\bm{x})\propto  p(\bm{x}|\bm{z},\bm{\theta}) p(\bm{z}|\bm{\theta})p(\bm{\theta}).
\end{equation}
In the context of the forward model and full-field analysis, Bayesian Hierarchical Models (BHMs) fall into this category.
A Bayesian forward approach involves using a given physical model to predict observations and then comparing these predictions with observations to infer the parameters of the model.
However, performing a global analysis on complex and large datasets, such as a lensing dataset, is not feasible. Therefore, the idea of dividing the global analysis into several subgroups, analyzing them in multiple steps, and using the output of each of those as input for the next one. From here, the \textit{hierarchical} nature of this methodology.
%--------------------------------------------------------------------
\paragraph{\textbf{Implicit inference}} referring to all the approaches that try to infer the marginal likelihood from simulations. This second class of approaches involves treating the simulator as a black box with only the ability to sample from the joint distribution:
\begin{equation}
    (\bm{x}, \bm{\theta})\sim p(\bm{x}, \bm{\theta}).
\end{equation}
Within this class of methods, we can differentiate between more traditional methods such as Approximate Bayesian Computation (ABC) and Density Estimation Likelihood-Free Inference (DELFI) methods. ABC employs rejection-criteria based approaches to approximate the likelihood by comparing simulations with data. 
In this work, our focus will be on DELFI methods, which approach the inference task as a density estimation problem. \\
In particular, the standard deep learning approach for Likelihood-Free Inference can be described as separated into two distinct steps:
\begin{enumerate}
    \item  Automatically learning an optimal low-dimensional summary statistic.
    \item Using Neural Density Estimation in low dimensions to infer the target distributions.
\end{enumerate}
In the first step, we introduce a parametric function $F_{\varphi}$ such that:
     \begin{equation}
         \bm{t}=F_{\varphi}(\bm{\theta}),
     \end{equation}
which aims to reduce the dimensionality of the data while preserving information. Typically, the compressed statistics $\bm{t}$ is assumed to have the same dimension of $\bm{\theta}$. \\
In the second step, Neural Density Estimation can target either building an estimate $p_{\varphi}$ of the likelihood function $p(\bm{x}|\bm{\theta})$ (referred to as the Neural Likelihood Estimation (NLE) task), or targeting the posterior distribution $p(\bm{\theta}|\bm{x})$, (known as Neural Posterior Estimation (NPE) task).

@@

Therefore, our primary focus will be to investigate the effectiveness and validity of these compression techniques in the context of weak-lensing applications (within the limits of our setting).

@@

%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{The SBILens framework}\label{Sec:the SBILens framework}
To investigate the questions above, we have developed the Python package \href{https://github.com/DifferentiableUniverseInitiative/sbi_lens}{\texttt{SBILens}}, a weak-lensing differentiable simulator based on a log-normal model. \texttt{SBILens} enables the sampling of convergence maps in a tomographic setting while considering the cross-correlation between different redshift bins.
\subsection{Lognormal Modeling}\label{Sec:Lognormal Modeling}
For various cosmological applications, the non-Gaussian field can be modeled as a Lognormal field \citep{coles1991lognormal,bohm2017bayesian}.
This model offers the advantage of generating the matter or convergence field rapidly while allowing the extraction of information beyond the two-point statistics. 
Although studies demonstrated that this model fails in describing the 3D field \citep{klypin2018density}, it properly describes the 2D convergence field \citep{clerkin2017testing, xavier2016improving}.
Assuming a simulated Gaussian convergence map $\kappa_g$, whose statistical properties are fully described by its power spectrum $C_{\ell}$ we know that this model is not a suitable representation of late-time and more evolved structures. One potential solution is to find a transformation $f(\kappa_g)$ of this map that captures the non-Gaussian features in the convergence field. In doing so, it is crucial to ensure that the transformed map maintains the correct mean and variance, effectively recovering the correct two-point statistics.
Denoting $\mu$ and $\sigma_g^2$ the mean and covariance matrix of $\kappa_g$ respectively, we can define the transformed convergence $\kappa_{ln}$ as a shifted lognormal random field:
\begin{equation}\label{Eq:log_norm_kappa}
    \kappa_{ln}=e^{\kappa_{g}}-\lambda, 
\end{equation}
where $\lambda$ is a free parameter that determines the shift of the lognormal distribution. The convergence $\kappa$ in a given redshift bin is fully determined by the shift parameter $\lambda$, the mean $\mu$ of the associated Gaussian field $\kappa_g$, and its variance $\sigma_{g}^2$.
The correlation of the lognormal field, denoted as $\xi_{ln}$, is also a function of these variables and is related to $\xi^{ij}_g$ through the following equations:
\begin{align}
    \xi^{ij}_{ln}(\theta) & \equiv \lambda_i \lambda_j (e^{ \xi^{ij}_g(\theta)}-1) \nonumber \\ 
    \xi^{ij}_g(\theta)&=\log{\left[ \frac{\xi^{ij}_{ln}(\theta)}{\lambda_i \lambda_j}+1\right ]}. \label{Eq:log_norm_corr}
\end{align}
Here $i$ and $j$ define a pair of redshift bins.
The parameter $\lambda$, also known as \textit{minimum convergence
parameter}, defines the lowest values for all possible values of $\kappa$.
The modeling of the shift parameter can be approached in various ways. For example, it can be determined by matching moments of the distribution \citep{xavier2016improving} or by treating it as a free parameter \citep{hilbert2011cosmic}. In general, the value of $\lambda$ depends on the redshift, cosmology, and the scale of the field at which smoothing is applied.

While it is straightforward to simulate a single map, if we want to constrain the convergence map in different redshift bins, an additional condition must be met. The covariance of the map should recover the correct angular power spectrum:
\begin{equation}\label{power_spectrum_definition}
    \left \langle \tilde{\kappa}^{(i)}_{ln} (\ell)\tilde{\kappa}^{*(j)}_{ln}(\ell')\right \rangle =C^{ij}_{ln}(\ell)\delta^{K}(\ell-\ell')
\end{equation}
where $ C^{ij}_{ln}(\ell)$ is the power spectrum of $\kappa_{ln}$ in Fourier space, defined as:
\begin{equation}\label{Eq:log_norm_cls}
    C^{ij}_{ln}(\ell)=2\pi \int_0^{\pi} d\theta \sin{\theta}P_{\ell}(\cos{\theta})\xi^{ij}_{ln}(\theta)
\end{equation}
and $P_{\ell}$ is the Legendre polynomial of order $\ell$. 
Using the lognormal model, we can simultaneously constrain the convergence field in different redshift bins while considering the correlation between the bins, as described by \autoref{Eq:log_norm_corr}.

@@

Each map is reproduced on a regular grid with dimensions of $256 \times 256$ pixels and covers an area of $10\times 10$ deg$^2$.  An example of a tomographic convergence map simulated using the \texttt{SBILens} package is shown in \autoref{fig:convergence_maps}. 

@@

    \includegraphics[width=\textwidth]{figures/example_kmaps.pdf}

@@

     Example of convergence maps simulated using the \texttt{SBILens} package.

@@

    Convergence power spectra for different tomographic bin combinations. The solid yellow line shows the measurement from 20 simulated maps using the survey setting described in Section \ref{Sec:the SBILens framework}, while the black dashed line shows the theoretical predictions computed using jax-cosmo. In this figure, the shaded regions represent the standard deviation from 20 independent map realizations.

@@

Additionally, we conduct a power spectrum study. Indeed, as discussed in Section \ref{Sec:Lognormal Modeling}, log-normal fields offer the advantage of rapidly generating convergent fields while accounting for non-Gaussianities. To emphasize this claim further, along with the full-field analysis, we include a power spectrum analysis. This analysis demonstrates that there is indeed a gain of information when adopting a full-field approach. \\

@@

To construct the explicit map-based inference strategy in the Hierarchical Bayesian framework, we built a likelihood based on the data model described in Section \ref{Sec:Lognormal Modeling}. 

@@

However, it is important to note that this approach does not guarantee the recovery of maximally informative summary statistics. Indeed, minimizing the $L_{2}$ norm is equivalent to training the model to estimate the mean of the posterior distribution.  We prove this statement in Section \ref{Sec:appendix_Mean Square Error}.

@@

In Section \ref{Sec:appendix_Maximum Absolute Error}, we demonstrate that minimizing this loss function is equivalent to training the model to estimate the median of the posterior distribution. While extensively employed in various previous studies \citep{2018PhRvD..97j3515G, fluri2018cosmological, ribli2019weak}, similar to the previous consideration for the MSE, it is important to note that the effectiveness of this loss function in extracting sufficient statistics needs to be demonstrated.

@@

An application of this method in the context of weak-lensing can be traced back to the work of \citet{fluri2021cosmological, fluri2022full}. Following \citet{charnock2018automatic}, they presented a compression approach that relies on the optimization of the Cramér-Rao bound. However, their approach does not assume that the summary follows a Gaussian likelihood. The loss function takes the form:
\begin{equation}\label{eq:loss_gfim}
\mathcal{L}=\log{\text{det}(\bm{\Sigma}_{\bm{\theta}}(\bm{t})})-2\log{\left|\text{det}\left(\frac{\partial \Psi_{\bm{\theta}}(\bm{t})}{\partial \bm{\theta}}\right)\right|},
\end{equation}
where $\bm{\Sigma}$ is the covariance matrix, and $\Psi_{\bm{\theta}}(\bm{t})=\mathbb{E}_{p(\bm {x}|\bm {\theta})}[\bm{t}]$.
This loss function is equivalent to the log-determinant of the inverse Fisher matrix used in \autoref{eq:loss_imnn}. \\
% \LM{TODO: condensed introduction; emphasize learning compression at single point; network learns score function which generalises smoothly away from fiducial \cite{Makinen_2021, Makinen_2022, Charnock_2018}, Wandelt (in prep.) 
% }
%--------------------------------------------------------------------
\subsubsection{Compression strategy}
%--------------------------------------------------------------------
The different compression strategies differ in the loss function employed during the training phase. For all of them, we use the same convolutional neural network architecture to build the compressor, specifically the ResNet-18, a deep ResNet architecture \citep{he2016deep}. The ResNet-18 is implemented using Haiku \citep{haiku2020github}, a Python deep learning library built on top of Jax.

@@

\subsection{Inference strategy}\label{Sec:Inference_strategy}

@@

        \caption{Table summarizing the different neural compression schemes used for weak-lensing applications. The table includes all studies conducted within the context of implicit analysis (Likelihood-free) and standard Likelihood-based analysis. \\

@@

We present the results of the constraints on the full $\Lambda$CDM parameter space expected for a survey like LSST-Y10. The results are obtained using the simulation procedure outlined in Section \ref{Sec:the SBILens framework} and the parameter inference strategy described in Section \ref{Sec:Inference_strategy}.

@@

We now compare the constraining power of the three different approaches described in Section \ref{Sec:experiment}: the standard two-point statistics and two map-based approaches, the explicit inference and the implicit inference strategy. 

@@

We note that the results obtained using different summaries are generally in agreement with each other, and there are no tensions present for any of the cosmological parameters. We report the marginalized summary constraints in \autoref{tab:summaries}. The results concern the cosmological parameters that are better constrained from weak-lensing: $\Omega_c$, $\sigma_8$, $w_0$.

@@

        Constraints on the $\Lambda$CDM parameter space as found in the LSST Y10 survey setup. The constraints are obtained by applying the $C_{\ell}$ (violet contours), the full field explicit inference (orange contours), and the full field implicit inference strategy (blue dashed contours), described in Section \ref{Sec:experiment}.

@@

        Constraints on the $\Lambda$CDM parameter space as found in the LSST Y10 survey setup. The constraints are obtained from three CNN map compressed statistics: the MSE (red contours), the MAE (black dashed contours), VMIM (yellow contours), the GNLL (blue dotted contours), described in Section \ref{Sec:experiment}.

@@

            $\Omega_c-  \sigma_8$ & 1222 & 2520 & 2526 & 2043 & 2316 & 1882 \\
            $\Omega_c-  w_0$      & 100  & 198  & 190  & 152  & 162  & 178  \\
            $\sigma_8-  w_0$      & 77   & 176  & 171  & 139  & 149  & 153  \\
